TADCompare: an R package for differential analysis and characterization of Topologically Associated Domains

### 2019-11-05 manuscript/manuscript.Rmd
Format for BMC Bioinformatics

**Results.** Using STRING and BioGrid PPI databases, we compared the coherence of 133 phenotype-associated gene sets from GWAScatalog against size-matched KEGG pathways (the reference for high coherence) and random networks (the lower limit of coherence). We observed a range of coherence estimates for each category of phenotypes. Metabolic traits and diseases were the most coherent, while psychiatric disorders and intelligence-related traits were the least coherent.

@@

We collected 133 phenotype-associated gene lists from the NHGRI-EBI GWAS catalog [@MacArthur:2017aa] (see Methods). They were grouped into seven disease categories (“Autoimmune”, “Cancer”, “Cardiovascular disease”, “Eye disease”, “Metabolic disease”, “Neurologic”, “Psychiatric”) and five trait categories (“Anthropometric trait”, “Cardiovascular trait”, “Eye trait”, “Intelligence”, “Metabolic trait”, Table 1, Supplementary Table 1).

@@

The coherence of phenotype-associated gene sets should be measured with respect to the realistic references of high and low coherence. To establish a reference of high coherence, we considered networks from collections of the Molecular Signatures Database (MSigDB v6.2) [@Subramanian:2005aa]. MSigDB contains sets of genes having various types of functional relationships. We found that pathways from the Kyoto Encyclopedia of Genes and Genomes (KEGG) [@Kanehisa:2000aa] had the smallest average slope corresponding to the highest level of coherence (Supplementary Table S2). Networks assembled from Gene Ontology - Cellular Component (GOCC) collection and REACTOME pathways followed, representing alternative sources of networks with high coherence. This is expected as genes expressed in the same cellular components (GOCC genes) or participating in the same metabolic pathways (REACTOME genes) are presumed to interact more frequently. In our study, KEGG networks were used in parallel with GOCC and REACTOME (referred hereafter as MSigDb networks) as a reference for high coherence. Conversely, as a reference for low coherence, we randomly sampled genes from a pool of genes in a given PPI database. These random gene lists represent expected network coherence arising by chance. The slopes of MSigDb and random networks represent reference levels for high and low coherence, respectively.

@@

+-------+
| ![](Figures/Fig1_network_distributions.png)      |
| **Figure 1. Degree distributions are affected by network size and configuration.** Degree distributions of A) ring networks and B) fully connected networks containing 10 and 20 nodes. While degree distributions of similarly coherent ring networks can be directly compared, degree distributions of fully connected networks depend on network size.   |
+-------+

To observe whether network properties affect coherence estimation in experimental settings, we investigated the effect of network size on the slopes of KEGG and random networks. We observed substantial negative correlation between network sizes and the slopes of KEGG and random networks (Pearson Correlation Coefficient (PCC) -0.58/-0.86, respectively, Figure 2, Supplementary Table S2). These observations confirm the notion that coherence estimation using internal vs. external degree distributions should be controlled for network size.

+-------+
| ![](Figures/Fig2_Size_Slope_Plot.png)      |
| **Figure 2. Network size is inversely associated with coherence.** Slopes of internal vs. external degree distributions (aka coherence) for KEGG and random networks are plotted against network size.   |
+-------+

@@

Given the dependence of degree distributions on network size, we designed a simple strategy to account for it. Briefly, for each phenotype-specific network of a given size, we created size-matched references of high and low coherence. Specifically, we selected 30 MSigDb networks (10 from KEGG, GOCC, and REACTOME collections) and 100 random networks matched in size to the corresponding network of phenotype-specific genes. Consequently, we use median slopes of internal vs. external degree distributions created by these size-matched references to normalize the slopes of phenotype-specific networks to the [0, 1] range (see Methods), where 0/1 correspond to random/high coherence, respectively. We found that the use of size-matched references indeed alleviates the dependency of coherence on network size (Average PCC = 0.01, Supplementary Figure S1) and allows for direct comparison of normalized coherences across phenotypes.

@@

In addition to size dependency, coherence estimates depend on the choice of the PPI database used to build networks and estimate the corresponding internal vs. external degree distributions. To evaluate the consistency of coherence estimates, we assessed the pairwise correlation of normalized coherences obtained using different PPI databases. Correlation between coherence estimates obtained using "STRING" and "STRING filtered" was high (PCC = 0.80, Supplementary Figure S1). Expectedly, coherence estimates using Biogrid database were less similar to that of "STRING" and "STRING filtered"; however, the overall correlation remained high. Notably, coherence estimates using "STRING filtered" database were most similar to that of "Biogrid" (PCC = 0.58), indicating that filtering step indeed removes "noisy" PPIs and improves quality of coherence estimates. Overall, these results demonstrate consistency in coherence estimates using different PPI databases.

@@

We estimated coherence of 49 traits from five categories (Figure 3, Supplementary Table S3, Supplementary Figures S2-S3). Among anthropometric traits, Weight and Body Mass Index had the highest level of coherence. Other examples of traits with high coherence include blood pressure-related traits (cardiovascular) and glucose-related traits (metabolic). Notably, the metabolic trait category had the highest overall coherence (mean coherence 0.86, Table 1), and these results were consistent when using other PPI databases (Supplementary Table S1). Several notable examples of low coherence include Obesity-related traits, Height (anthropometric traits), and blood count-related traits (cardiovascular). Traits in the Intelligence category had low overall coherence (mean coherence = 0.41).

+-------+
| ![](Figures/Fig3_summary_traits.png)      |
| **Figure 3. Coherence estimates of traits.** Size of dots represents the level of normalized coherence (X-axis) for individual traits (Y-axis).   |
+-------+

@@

We further estimated coherence of 67 diseases from seven categories (Figure 4, Supplementary Figures S4-S5, Supplementary Table S3). Metabolic diseases had high overall coherence (average coherence 0.91, Table 1), with "Obesity" and "Type 2 diabetes" being the most coherent (1.01 and 0.91, respectively). Similarly, the average coherence for cancer diseases was 0.72, with "Testicular germ cell tumor" and "Pancreatic cancer" being highly coherent (0.92 and 0.74, respectively).

@@

"Rheumatoid arthritis", "Systemic Lupus Erythematosus", and "Systemic sclerosis" were the most coherent in the "Autoimmune" category (0.86, 0.80, 0.79, respectively). On the other end of the coherence spectrum, "Allergic rinitis" and "Atopic dermatitis" were the least coherent (0.51, 0.60, respectively). The average coherence of autoimmune diseases was high, 0.70 (Table 1), with all autoimmune diseases, except "Allergic rinitis", having coherence estimates > 0.60 (Supplementary Table S3). Diseases in the "Neurologic" category were similarly highly coherent (average coherence 0.72). However, the average coherence of neurologic disorders using STRING database was 0.49, suggesting that, with larger number of networks being estimated the overall coherence of neurologic diseases is low. "Alzheimer's disease" and "Parkinson's disease" were the most coherent (0.88 and 0.84, respectively), while "Migraine" was the least coherent (0.54). Diseases in the "Eye disease" category were more heterogeneous, with coherence ranging from 0.35 for "Corneal astigmatism" to 1.01 for "Refractive error". The average coherence of diseases in the "Eye disease" category was relatively high (0.66).

+-------+
| ![](Figures/Fig4_summary_diseases.png)      |
| **Figure 4. Coherence estimates of diseases.** Size of dots represents the level of normalized coherence (X-axis) for individual diseases (Y-axis).   |
+-------+

@@

Given the wide range of phenotype-specific coherence estimates, a natural question is whether they are significantly larger than the coherence of random networks. We assessed the significance of coherence estimates using a permutation test (see Methods), individually for each PPI database. Permutation p-values obtained using Biogrid and STRING filtered databases were highly correlated (PCC = 0.64, Supplementary Figure S1), as would be expected for high-quality PPIs. Permutation p-values using STRING databases were predominantly significant (p-value < 0.001) due to a large number of PPIs that enabled phenotype-specific networks to be consistently different from random. Permutation p-values inversely correlated with coherence estimates (average PCC = -0.23), as would be expected for less coherent networks that resemble random networks. This dependency was especially pronounced for database-specific estimates, e.g., Biogrid coherence and p-value estimates correlated with PCC = -0.69 (STRING filtered PCC = -0.53, Supplementary Figure S1). These observations confirmed that coherence and p-value estimates remained consistent when using different PPI databases.

The majority of phenotype-specific networks (~85%) were significantly more coherent than random networks (STRING filtered p-value < 0.001, Supplementary Table S3). Among less significant phenotypes were "White blood cell count" (p-value = 0.085) and other blood count-related phenotypes, "Corneal astigmatism" (p-value = 0.04), "Major depressive disorder" (p-value = 0.03). When using Biogrid databases, "Educational attainment", "Myopia (pathological)", and "Celiac disease" were among the non-significant phenotypes (p-values 0.30, 0.20, 0.16, respectively). Interestingly, "Schizophrenia" network, despite being among the least coherent, was consistently significantly different from random (p-value < 0.0001). Overall, the significance estimation of coherence supports our observation that phenotype-specific networks in "Intelligence" and "Psychiatric disease" categories generally have low coherence that is frequently statistically insignificant. In summary, these results confirm our ability to identify coherence of phenotype-specific networks and test their significance over random networks.

@@

Although using median slopes of MSigDb and random networks is expected to be a robust reference for high/random coherence, some phenotype-specific networks may have coherence larger than 1, that is, being more coherent than the median coherence of MSigDb networks. These situations may occur in smaller networks that have a higher chance to contain genes with unusually high internal degree distributions. Such outliers would inflate coherence estimations. Similar outliers may be observed in larger networks. For these reasons, coherence estimates for excessively large networks ("Obesity-related traits", 779 genes, "Height", 418 genes) were excluded from calculating the correlation between coherences obtained using different PPI databases (Supplementary Figure S1).

@@

# Supplementary material

@@

**Supplementary Figure S5. Coherence estimates of diseases using STRING as a reference PPI database.** See legend for Supplementary Figure S2.

>>>>

**Results.** Using STRING and BioGrid PPI databases, we compared the coherence of 116 phenotype-associated gene sets from GWAScatalog against size-matched KEGG pathways (the reference for high coherence) and random networks (the lower limit of coherence). We observed a range of coherence estimates for each category of phenotypes. Metabolic traits and diseases were the most coherent, while psychiatric disorders and intelligence-related traits were the least coherent.

@@

We collected 133 phenotype-associated gene lists from the NHGRI-EBI GWAS catalog [@MacArthur:2017aa] (see Methods). They were grouped into seven disease categories (“Autoimmune”, “Cancer”, “Cardiovascular disease”, “Eye disease”, “Metabolic disease”, “Neurologic”, “Psychiatric”) and five trait categories (“Anthropometric trait”, “Cardiovascular trait”, “Eye trait”, “Intelligence”, “Metabolic trait”, Table 1, Additional File 1).

@@

The coherence of phenotype-associated gene sets should be measured with respect to the realistic references of high and low coherence. To establish a reference of high coherence, we considered networks from collections of the Molecular Signatures Database (MSigDB v6.2) [@Subramanian:2005aa]. MSigDB contains sets of genes having various types of functional relationships. We found that pathways from the Kyoto Encyclopedia of Genes and Genomes (KEGG) [@Kanehisa:2000aa] had the smallest average slope corresponding to the highest level of coherence (Additional File 2). Networks assembled from Gene Ontology - Cellular Component (GOCC) collection and REACTOME pathways followed, representing alternative sources of networks with high coherence. This is expected as genes expressed in the same cellular components (GOCC genes) or participating in the same metabolic pathways (REACTOME genes) are presumed to interact more frequently. In our study, KEGG networks were used in parallel with GOCC and REACTOME (referred hereafter as MSigDb networks) as a reference for high coherence. Conversely, as a reference for low coherence, we randomly sampled genes from a pool of genes in a given PPI database. These random gene lists represent expected network coherence arising by chance. The slopes of MSigDb and random networks represent reference levels for high and low coherence, respectively.

@@

<!--
+-------+
| ![](Figures/Fig1_network_distributions.png)      |
| **Figure 1. Degree distributions are affected by network size and configuration.** Degree distributions of A) ring networks and B) fully connected networks containing 10 and 20 nodes. While degree distributions of similarly coherent ring networks can be directly compared, degree distributions of fully connected networks depend on network size.   |
+-------+
-->
**Figure 1. Degree distributions are affected by network size and configuration.** Degree distributions of A) ring networks and B) fully connected networks containing 10 and 20 nodes. While degree distributions of similarly coherent ring networks can be directly compared, degree distributions of fully connected networks depend on network size.

To observe whether network properties affect coherence estimation in experimental settings, we investigated the effect of network size on the slopes of KEGG and random networks. We observed substantial negative correlation between network sizes and the slopes of KEGG and random networks (Pearson Correlation Coefficient (PCC) -0.58/-0.86, respectively, Figure 2, Additional File 2). These observations confirm the notion that coherence estimation using internal vs. external degree distributions should be controlled for network size.

<!--
+-------+
| ![](Figures/Fig2_Size_Slope_Plot.png)      |
| **Figure 2. Network size is inversely associated with coherence.** Slopes of internal vs. external degree distributions (aka coherence) for KEGG and random networks are plotted against network size.   |
+-------+
-->
**Figure 2. Network size is inversely associated with coherence.** Slopes of internal vs. external degree distributions (aka coherence) for KEGG and random networks are plotted against network size.

@@

Given the dependence of degree distributions on network size, we designed a simple strategy to account for it. Briefly, for each phenotype-specific network of a given size, we created size-matched references of high and low coherence. Specifically, we selected 30 MSigDb networks (10 from KEGG, GOCC, and REACTOME collections) and 100 random networks matched in size to the corresponding network of phenotype-specific genes. Consequently, we use median slopes of internal vs. external degree distributions created by these size-matched references to normalize the slopes of phenotype-specific networks to the [0, 1] range (see Methods), where 0/1 correspond to random/high coherence, respectively. We found that the use of size-matched references indeed alleviates the dependency of coherence on network size (Average PCC = 0.01, Additional File 4) and allows for direct comparison of normalized coherences across phenotypes.

@@

In addition to size dependency, coherence estimates depend on the choice of the PPI database used to build networks and estimate the corresponding internal vs. external degree distributions. To evaluate the consistency of coherence estimates, we assessed the pairwise correlation of normalized coherences obtained using different PPI databases. Correlation between coherence estimates obtained using "STRING" and "STRING filtered" was high (PCC = 0.80, Additional File 4). Expectedly, coherence estimates using Biogrid database were less similar to that of "STRING" and "STRING filtered"; however, the overall correlation remained high. Notably, coherence estimates using "STRING filtered" database were most similar to that of "Biogrid" (PCC = 0.58), indicating that filtering step indeed removes "noisy" PPIs and improves quality of coherence estimates. Overall, these results demonstrate consistency in coherence estimates using different PPI databases.

@@

We estimated coherence of 49 traits from five categories (Figure 3, Additional File 3, Additional File 5, 6). Among anthropometric traits, Weight and Body Mass Index had the highest level of coherence. Other examples of traits with high coherence include blood pressure-related traits (cardiovascular) and glucose-related traits (metabolic). Notably, the metabolic trait category had the highest overall coherence (mean coherence 0.86, Table 1), and these results were consistent when using other PPI databases (Additional File 1). Several notable examples of low coherence include Obesity-related traits, Height (anthropometric traits), and blood count-related traits (cardiovascular). Traits in the Intelligence category had low overall coherence (mean coherence = 0.41).

<!--
+-------+
| ![](Figures/Fig3_summary_traits.png)      |
| **Figure 3. Coherence estimates of traits.** Size of dots represents the level of normalized coherence (X-axis) for individual traits (Y-axis).   |
+-------+
-->
**Figure 3. Coherence estimates of traits.** Size of dots represents the level of normalized coherence (X-axis) for individual traits (Y-axis).

@@

We further estimated coherence of 67 diseases from seven categories (Figure 4, Additional File 7, 8, Additional File 3). Metabolic diseases had high overall coherence (average coherence 0.91, Table 1), with "Obesity" and "Type 2 diabetes" being the most coherent (1.01 and 0.91, respectively). Similarly, the average coherence for cancer diseases was 0.72, with "Testicular germ cell tumor" and "Pancreatic cancer" being highly coherent (0.92 and 0.74, respectively).

@@

"Rheumatoid arthritis", "Systemic Lupus Erythematosus", and "Systemic sclerosis" were the most coherent in the "Autoimmune" category (0.86, 0.80, 0.79, respectively). On the other end of the coherence spectrum, "Allergic rinitis" and "Atopic dermatitis" were the least coherent (0.51, 0.60, respectively). The average coherence of autoimmune diseases was high, 0.70 (Table 1), with all autoimmune diseases, except "Allergic rinitis", having coherence estimates > 0.60 (Additional File 3). Diseases in the "Neurologic" category were similarly highly coherent (average coherence 0.72). However, the average coherence of neurologic disorders using STRING database was 0.49, suggesting that, with larger number of networks being estimated the overall coherence of neurologic diseases is low. "Alzheimer's disease" and "Parkinson's disease" were the most coherent (0.88 and 0.84, respectively), while "Migraine" was the least coherent (0.54). Diseases in the "Eye disease" category were more heterogeneous, with coherence ranging from 0.35 for "Corneal astigmatism" to 1.01 for "Refractive error". The average coherence of diseases in the "Eye disease" category was relatively high (0.66).

<!--
+-------+
| ![](Figures/Fig4_summary_diseases.png)      |
| **Figure 4. Coherence estimates of diseases.** Size of dots represents the level of normalized coherence (X-axis) for individual diseases (Y-axis).   |
+-------+
-->
**Figure 4. Coherence estimates of diseases.** Size of dots represents the level of normalized coherence (X-axis) for individual diseases (Y-axis).

@@

Given the wide range of phenotype-specific coherence estimates, a natural question is whether they are significantly larger than the coherence of random networks. We assessed the significance of coherence estimates using a permutation test (see Methods), individually for each PPI database. Permutation p-values obtained using Biogrid and STRING filtered databases were highly correlated (PCC = 0.64, Additional File 4), as would be expected for high-quality PPIs. Permutation p-values using STRING databases were predominantly significant (p-value < 0.001) due to a large number of PPIs that enabled phenotype-specific networks to be consistently different from random. Permutation p-values inversely correlated with coherence estimates (average PCC = -0.23), as would be expected for less coherent networks that resemble random networks. This dependency was especially pronounced for database-specific estimates, e.g., Biogrid coherence and p-value estimates correlated with PCC = -0.69 (STRING filtered PCC = -0.53, Additional File 4). These observations confirmed that coherence and p-value estimates remained consistent when using different PPI databases.

The majority of phenotype-specific networks (~85%) were significantly more coherent than random networks (STRING filtered p-value < 0.001, Additional File 3). Among less significant phenotypes were "White blood cell count" (p-value = 0.085) and other blood count-related phenotypes, "Corneal astigmatism" (p-value = 0.04), "Major depressive disorder" (p-value = 0.03). When using Biogrid databases, "Educational attainment", "Myopia (pathological)", and "Celiac disease" were among the non-significant phenotypes (p-values 0.30, 0.20, 0.16, respectively). Interestingly, "Schizophrenia" network, despite being among the least coherent, was consistently significantly different from random (p-value < 0.0001). Overall, the significance estimation of coherence supports our observation that phenotype-specific networks in "Intelligence" and "Psychiatric disease" categories generally have low coherence that is frequently statistically insignificant. In summary, these results confirm our ability to identify coherence of phenotype-specific networks and test their significance over random networks.

@@

Although using median slopes of MSigDb and random networks is expected to be a robust reference for high/random coherence, some phenotype-specific networks may have coherence larger than 1, that is, being more coherent than the median coherence of MSigDb networks. These situations may occur in smaller networks that have a higher chance to contain genes with unusually high internal degree distributions. Such outliers would inflate coherence estimations. Similar outliers may be observed in larger networks. For these reasons, coherence estimates for excessively large networks ("Obesity-related traits", 779 genes, "Height", 418 genes) were excluded from calculating the correlation between coherences obtained using different PPI databases (Additional File 4).

@@

# Additional files

**Additional File 1. Summary statistics of the analyzed phenotype categories.** Minimum, mean, median, maximum of coherence estimates, the number of genes, SNPs, the total number of phenotype networks. Each worksheet corresponds to the PPI database used.

**Additional File 2. Summary of networks from MSigDB categories.** Minimum, mean, median, maximum of the slopes, the number of genes, and the total number of networks are shown for each collection, along with the correlation between network size and slope. "CP:/KEGG/Reactome/Biocarta" - canonical pathways, "GOBP/GOMF/GOCC" - gene ontology biological processes, molecular functions, cellular component, "TFT" - transcription factor targets, "CGN" - cancer gene neighborhoods, "CM" - cancer modules, "MIR" - microRNA targets, "CGP" - chemical and genetic perturbations.

**Additional File 3. Phenotype-specific coherence estimates.** The PPI database (STRING, STRING filtered, Biogrid) are specified in the corresponding column names. For each phenotype category, the results are sorted by the "String Filt Coherence" column in descending order. "NA" indicates the corresponding value cannot be estimated due to lack of sufficient number of genes annotated with PPIs.

**Additional File 4. Normalization of coherence estimates alleviates its network size dependence.** Red/blue gradient and numbers represent Pearson correlation coefficients for each pairwise comparison of the number of SNPs, genes, and the normalized coherence estimates using the corresponding PPI databases.

**Additional File 5. Coherence estimates of traits using Biogrid as a reference PPI database.** Size of dots represents the level of normalized coherence (X-axis) for individual traits (Y-axis). Plots are faceted by categories. Missing entries indicate that, for a given trait, a network could not be built and the coherence cannot be estimated.

**Additional File 6. Coherence estimates of traits using STRING as a reference of PPI database.** See legend for Additional File 5.

**Additional File 7. Coherence estimates of diseases using Biogrid as a reference PPI database.** See legend for Additional File 5.

**Additional File 8. Coherence estimates of diseases using STRING as a reference PPI database.** See legend for Additional File 5.

<!--
# Supplementary material

@@

**Supplementary Figure S5. Coherence estimates of diseases using STRING as a reference PPI database.** See legend for Supplementary Figure S2.
-->
